Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-25 main.tex
Edits to methods down to statistics.

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell that was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
(Bonte Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde's design include simplifying the
Bump'Em motor controller with an inexpensive microcontroller and the use of a
simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The participants were all confident in their cycling skills and
had cycled at least once in the last month. All participants consented to the
experiment and could decline to continue at any time. The study was approved by
Delft University of Technology's Human Research Ethics Council (\#3897).

The participants were divided into two groups. The first group of fifteen
participants
performed the protocol at a belt speed of
10~\si{\kilo\meter\per\hour}(2.8~\si{\meter\per\second}) with the controller
gain set to $g=8$ and the second group of eleven particpants performed the
protocol at a belt speed of 6~\si{\kilo\meter\per\hour}
(1.7~\si{\meter\per\second}) with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and they were attached to the ceiling via a fall safety
harness, Figure~\ref{fig:participant-in-set-up}. We allowed them to practice
riding on the treadmill until they indicated they were comfortable enough to
have perturbations applied. For most, this was less than a 10~\si{\minute} warm
up.  We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill
as a final preparation before the perturbations.  We then applied perturbations
in random directions (clockwise or counter-clockwise), starting at
20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
participants fell. We define a ``fall'' on the treadmill by two criteria: 1)
the rider removes their foot from the pedal and places it on the ground or 2)
the bicycle wheel exceeds the width of the treadmill belt.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We log the magnitude that causes the first fall to characterize
that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is determined
at random. If the perturbation results in a fall, the estimate of the
perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
perturbation did not result in a fall. Five new possible perturbation forces
are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of perturbations for each participant.

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
compared to our Bump'Em controller command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Externally applied perturbation torque alongside the resulting
    measured motion and steer motor induced torque based on a 110~\si{\newton}
    counterclockwise applied force at a 6~\si{\kph} travel speed. The circles
    on the roll and steer angle plot show the angles \(\phi_0\) and
    \(\delta_0\) at the perturbation start.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
  \label{eq:angular-impulse}
\end{align}

where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}.

We record the order number of each perturbation, i.e. first, second, third,
\ldots, to measure how many perturbations a rider is subjected to before the
current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

>>>>

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on a microprocessor (Arduino Mega 2560,
Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors are
arranged at the four corners of a 1~\si{\meter} wide by 2~\si{\meter} long
treadmill (Bonte Technology B.V., Zwolle, The Netherlands) that can reach
speeds up to 18~\si{\kilo\meter\per\hour}. The general design of this
perturbation system is described in detail in Van De Velde's MSc
thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
Velde's design include simplifying the Bump'Em motor controller with an
inexpensive microcontroller and the use of a simpler non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The participants were all confident in their cycling skills
and had cycled at least once in the last month. All participants consented to
the experiment and could decline to continue at any time. The study was
approved by Delft University of Technology's Human Research Ethics Council
(record \#3897).

The participants were divided into two groups. The first group of fifteen
participants performed the protocol at a belt speed of
10~\si{\kph}(2.8~\si{\mps}) with the controller gain value set to $g=8$ and the
second group of eleven particpants performed the protocol at a belt speed of
6~\si{\kph} (1.7~\si{\mps}) with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the 6~\si{\kph}
group.

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
free movement pre-fall. The participants practiced riding on the treadmill
until they indicated they were comfortable enough to have perturbations
applied. For most, this was less than a 10~\si{\minute} warm up. We then asked
the rider to ride for 90~\si{\second}, while attempting to maintain the
location of their front wheel on the center line of the treadmill as a baseline
measure before the perturbations. We then applied perturbations in random
directions (clockwise or counter-clockwise), starting at 20~\si{\newton} and
increasing the magnitude by 30~\si{\newton} until the participants fell. We
defined a ``fall'' on the treadmill by two criteria: 1) the rider removes their
foot from the pedal and places it on the ground or 2) the bicycle wheel exceeds
the width of the treadmill belt. Figure~\ref{fig:perturbation-sequence} shows
an example resulting motion from a perturbation. We logged the force magnitude
that caused the first fall to characterize that participant's
\emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen first is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of 20 perturbations for each participant.

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle \(\delta\), roll angle \(\phi\), roll angular rate
\(\dot{\phi}\), and forward speed \(v\). Figure~\ref{fig:perturbation_0} shows
an example perturbation force measurement compared to our Bump'Em controller
command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Externally applied perturbation torque alongside the resulting
    measured motion and steer motor induced torque based on a 110~\si{\newton}
    counterclockwise applied force at a 6~\si{\kph} travel speed. The circles
    on the roll and steer angle plot show the angles \(\phi_0\) and
    \(\delta_0\) at the perturbation start.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance assist in
\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
  \label{eq:angular-impulse}
\end{align}

where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}
